---
title: "RLadies 2019"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
---

## Introduction
This notebook provides an overview of some of the features and tools that are at the core of the `tidystl` framework. More about `tidystl` can be found on the [SLU openGIS website](https://slu-opengis.github.io/news/tidystl_annoucement/). 

## Dependencies
This notebook requires a variety of `tidystl`, `tidyverse`, and other dependencies. These reflect the ways in which I manage my own research and teach students to use `R`. There are other frameworks both for data wrangling (i.e. base `R`, `dt` (datatable)) and spatial data management (i.e. `sp`).

```{r load-packages}
# tidystl packages
library(compstatr)     # stlmpd data wrangling
library(gateway)       # st louis spatial data tools
library(postmastr)     # street address normalization
library(stlcsb)        # st louis citizens service bureau data wrangling

# tidyverse packages
library(dplyr)         # data wrangling
library(ggplot2)       # plotting

# spatial packages
library(sf)            # spatial data tools

# other packages
library(here)          # file path management
```

## Load Data
This notebook requires a number of data sets, some of which we can load right away. Two local geocoders are saved as `.rda` files in the `data/` to provide the most efficient storage means possible:

```{r load-geocoders}
# load local full geocoder
load(here("data", "geocoder.rda"))

# load local short geocoder
load(here("data", "geocoder_s.rda"))
```

Both geocoders could also be built using the `gateway` package:

```r
# build local full geocoder
geocoder <- gw_build_geocoder(style = "full", class = "tibble", return = "coords")

# build local short goecoder
geocoder_s <- gw_build_geocoder(style = "short", class = "tibble", return = "coords")
```

These are not built here to save time during the tutorial.
